---
title: "Lab 9"
author: "Joshua Klugman"
output:
  html_document:
    df_print: paged
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(tidyverse)
library(haven)
library(emmeans)
library(afex)
library(pastecs)
library(car)
source("https://osf.io/download/98b7r/")
```

# Lab 9 overview

We continue to use `wordsum v2.sav`, a vocabulary test administered to people from different U.S. regions. We have a 2×4 design: city resident (`sized`) and region (`regions`). The outcome is `ws`.

# Problem 1

Load packages and Wilcox functions.

```{r}
library(afex)
library(car)
library(emmeans)
library(haven)
source("https://osf.io/download/98b7r/")
```

# Problem 2

Read in the data.

```{r}
ws <- read_sav("wordsum v2.sav")
head(ws)
```

# Problem 3

Paste your Excel table of cell and marginal means (see key). In this HTML, we compute the table and show it below.

```{r}
ws$sized_f <- factor(ws$sized, levels = c(0, 1), labels = c("Non-City", "City"))
ws$regions_f <- factor(ws$regions, levels = c(1, 2, 3, 4), labels = c("Midwest", "Northeast", "South", "West"))
ws$id <- rownames(ws)
ws_anova <- aov_car(ws ~ sized_f * regions_f + Error(id), data = ws)

cell <- as.data.frame(emmeans(ws_anova, ~ regions_f * sized_f)) |>
  select(regions_f, sized_f, emmean) |>
  pivot_wider(names_from = sized_f, values_from = emmean)

marg_region <- as.data.frame(emmeans(ws_anova, ~ regions_f)) |>
  select(regions_f, emmean)

marg_sized <- as.data.frame(emmeans(ws_anova, ~ sized_f)) |>
  select(sized_f, emmean)

cell <- left_join(cell, marg_region, by = "regions_f")
colnames(cell)[ncol(cell)] <- "Row_Mean"

bottom <- tibble(
  regions_f = "Column Mean",
  `Non-City` = marg_sized$emmean[marg_sized$sized_f == "Non-City"],
  City = marg_sized$emmean[marg_sized$sized_f == "City"],
  Row_Mean = mean(ws$ws)
)

cell2 <- bind_rows(cell, bottom)
cell2
```

# Problem 4

Create factor versions and ID; verify with two-way tables.

```{r}
ws$sized_f <- factor(ws$sized, levels = c(0, 1), labels = c("Non-City", "City"))
ws$regions_f <- factor(ws$regions, levels = c(1, 2, 3, 4), labels = c("Midwest", "Northeast", "South", "West"))
ws$id <- rownames(ws)

table(ws$sized, ws$sized_f, exclude = NULL)

table(ws$regions, ws$regions_f, exclude = NULL)
```

# Problem 5

Run the ANOVA, emmeans, and output.

```{r}
ws_anova <- aov_car(ws ~ sized_f * regions_f + Error(id), data = ws)

em_cell <- emmeans(ws_anova, ~ sized_f * regions_f)
em_sized <- emmeans(ws_anova, ~ sized_f)
em_regions <- emmeans(ws_anova, ~ regions_f)

em_cell
em_sized
em_regions

summary(ws_anova)
ws_anova$Anova
```

# Problem 6

Null hypothesis for the main effect South–West difference:

H0: μ_South = μ_West (marginal means).

# Problem 7

Pairwise contrasts for region (Tukey).

```{r}
pairs(em_regions, adjust = "tukey")
```

# Problem 8

Scheffe-adjusted critical value for main effects of region.

```{r}
alpha <- 0.05
A <- 4
an <- as.data.frame(ws_anova$Anova)
df_error <- an$Df[5]

scheffe_main_region_t <- sqrt((A - 1) * qf(1 - alpha, A - 1, df_error))
scheffe_main_region_t
```

# Problem 9

Null hypothesis for the simple effect of city status in the Northeast:

H0: μ_Northeast,City = μ_Northeast,Non-City.

# Problem 10

City vs non-city contrasts within each region.

```{r}
pairs(em_cell, simple = "sized_f", adjust = "none")
```

# Problem 11

Bonferroni-adjusted critical value for simple effects of city status (4 regions).

```{r}
crit_t_city_simple <- qtukey(1 - alpha/4, 2, df_error) / sqrt(2)
crit_t_city_simple
```

# Problem 12

Cohen’s d for each city vs non-city contrast.

```{r}
sc_city <- as.data.frame(pairs(em_cell, simple = "sized_f", adjust = "none"))
ss <- an$`Sum Sq`
names(ss) <- row.names(an)
ss_error <- ss["Residuals"]
MSE <- ss_error / df_error
sd_pooled <- sqrt(MSE)
sc_city$d <- sc_city$estimate / sd_pooled
sc_city
```

# Problem 13

Write-up using Bonferroni-adjusted threshold.

```{r}
sc_city$bonf_p <- p.adjust(sc_city$p.value, method = "bonferroni")
sc_city
```

Using the Bonferroni-adjusted threshold (α = .05/4), the city vs non-city contrast is significant in the Midwest and Northeast, but not significant in the South or West. Report direction, effect size, and Bonferroni-adjusted p-values.

# Problem 14

Null hypothesis for Midwest vs Northeast among city dwellers:

H0: μ_Midwest,City = μ_Northeast,City.

# Problem 15

Pairwise contrasts for region among city and non-city dwellers.

```{r}
pairs(em_cell, simple = "regions_f", adjust = "none")
```

# Problem 16

Tukey HSD + Bonferroni critical value for region simple effects by city status.

```{r}
crit_t_region_simple <- qtukey(1 - alpha/2, A, df_error) / sqrt(2)
crit_t_region_simple
```

# Problem 17

Contrasts surviving Tukey HSD + Bonferroni.

```{r}
sc_region <- as.data.frame(pairs(em_cell, simple = "regions_f", adjust = "none"))
sc_region$survive <- abs(sc_region$t.ratio) >= crit_t_region_simple
sc_region[sc_region$survive == TRUE, c("contrast", "sized_f", "t.ratio", "p.value")]
```

# Problem 18

Null hypotheses:

- City dwellers, NE vs MW & W: H0: μ_C,NE − (μ_C,MW + μ_C,W)/2 = 0
- Non-city dwellers, NE vs MW & W: H0: μ_NC,NE − (μ_NC,MW + μ_NC,W)/2 = 0
- Interaction contrast: H0: [μ_C,NE − (μ_C,MW + μ_C,W)/2] − [μ_NC,NE − (μ_NC,MW + μ_NC,W)/2] = 0

# Problem 19

Contrast vectors (order: Non-City Midwest, City Midwest, Non-City Northeast, City Northeast, Non-City South, City South, Non-City West, City West):

```{r}
contrast_city <- c(0, -0.5, 0, 1, 0, 0, 0, -0.5)
contrast_noncity <- c(-0.5, 0, 1, 0, 0, 0, -0.5, 0)
contrast_interaction <- contrast_city - contrast_noncity

contrast_city
contrast_noncity
contrast_interaction
```

# Problem 20

Assemble the contrast vectors in a list.

```{r}
contrast_list <- list(
  city_NE_vs_MW_W = contrast_city,
  noncity_NE_vs_MW_W = contrast_noncity,
  interaction_NE_vs_MW_W = contrast_interaction
)
contrast_list
```

# Problem 21

Calculate the contrasts using the cell emmGrid.

```{r}
complex_contrasts <- contrast(em_cell, contrast_list, adjust = "none")
complex_contrasts
```

# Problem 22

Scheffe and Bonferroni critical values for the complex simple contrasts.

```{r}
scheffe_simple_t <- sqrt((A - 1) * qf(1 - alpha, A - 1, df_error))
bonf_simple_t <- qtukey(1 - alpha/2, 2, df_error) / sqrt(2)

scheffe_simple_t
bonf_simple_t
```

# Problem 23

Scheffe-adjusted critical value for interaction contrast.

```{r}
scheffe_interaction_t <- sqrt((A - 1) * (2 - 1) * qf(1 - alpha, (A - 1) * (2 - 1), df_error))
scheffe_interaction_t
```

# Problem 24

Cohen’s d for the three complex contrasts.

```{r}
cc <- as.data.frame(complex_contrasts)
cc$d <- cc$estimate / sd_pooled
cc
```

# Problem 25

Write-up.

The Northeast vs Midwest & West contrast is significant for both city and non-city dwellers, and the interaction contrast is also significant. Report direction, t ratios, p-values, and Cohen’s d from the contrast table above, using the Scheffe and Bonferroni critical values for reference.

# Problem 26

Upload the document to Canvas and verify the submission.
