---
title: "Lab Session 24 ANSWER KEY"
author: "Joshua Klugman"
output: 
  word_document:
    highlight: kate
    reference_docx: style_template_ak2.docx
    fig_caption: no
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# The Polynomial Specification

Using data from the 1991 General Social Survey (GSS), we are going to use a polynomial regression function to examine the association between schooling (independent variable; IV) and personal income in $10,000s (dependent variable; DV).

## Problems 1-7 are a repeat of the prior lab session.

## 1. Load packages and set options

```{r, eval=F}
library(tidyverse)
library(psych)
library(stargazer)
library(haven)
library(effects)
options(digits=6)
```


## 2. Load data

```{r, eval=F}
df_raw <- haven::read_sav("GSS1991.sav")
```


## 3. Select only variables of interest.

Let’s make the data a bit friendlier to work with by selecting only a few variables of interest, `RINCOM91`, `CONRINC`, and `EDUC` (`RINCOM91` is the original ordinal coded version of personal income; `CONRINC` used midpoint coding to get a  continuous version.  We will use `RINCOM91` only for data management and not for any analyses).

```{r, eval=F}
df<-subset(df_raw, select=c(RINCOM91, CONRINC, EDUC))
```


## 4. Get value labels and distributions for all variables.

```{r, eval=T}
attr(df$EDUC, "labels")
table(df$EDUC, exclude = NULL)

attr(df$CONRINC, "labels")
table(df$CONRINC, exclude = NULL)


attr(df$RINCOM91, "labels")
table(df$RINCOM91, exclude = NULL)

```



## 5. Coerce all variables to be numeric

```{r, eval=F}
df$RINCOM91<-as.numeric(df$RINCOM91)
df$CONRINC<-as.numeric(df$CONRINC)
df$EDUC<-as.numeric(df$EDUC)
```

## 6. Make sure there are no numeric missing values 

E.g. 0, 999998 or 999999 for "Inapplicable", "Don't know", "No answer" responses for CONRINC.  We want all missing values to be *system* missing values indicated as `<NA>`.

### Answer

All missing values should be system missing.


## 7. Subset the data to just include cases with complete information on both variables


```{r, eval=T}
df_ss<-na.omit(df)
```

## At this point, we are diverging from the prior lab.  

## 8. Calculate income in $10,000s

Store this as a new variable (`income10k`) in the subsetted data frame.  From here on out, this is the income variable you will use for this lab session.

### Answer

```{r}
df_ss$income10k<-df_ss$CONRINC/10000
```

## 9. Fit a polynomial regression model

There are two ways to do this, pick your favorite! Use summary() to get model output.


```{r, eval=T}
m1 <- lm(data = df_ss,
         income10k ~ poly(EDUC, degree = 2, raw = TRUE))
#In m1, we use lm() to regress income (in $10,000s) on education as a second-degree raw polynomial, meaning poly() automatically generates both the linear and squared terms for us in one call.
m2 <- lm(data = df_ss,
         income10k ~ EDUC + I(EDUC^2))
#In m2, we do the exact same thing manually — we regress income on education and education-squared, using I(EDUC^2). note: we  to use I() here to tell R that the caret means "square this" rather than its usual formula meaning of interaction order.

summary(m1)
```
*A quadratic regression of income (in $10k) on education revealed a significant curvilinear relationship, with both the linear (b = -0.43, p = .019) and quadratic (b = 0.03, p < .001) terms significant, suggesting that the positive effect of education on income accelerates at higher levels of education. The model accounted for approximately 14.8% of the variance in income (R² = .148, F(2, 908) = 79, p < .001).*

Use psych::describe() to get more information about our variables.

```{r, eval=F}
psych::describe(df_ss,quant=c(.1,.9))
#Adding quant=c(.1,.9) tells R to also return the 10th and 90th percentiles, which gives you a quick sense of the spread of the middle 80% of your data
```

## 10. Create a `stargazer` table

Remember to check automatically generated variables before adding your own covariate and dependent variable labels.  

Note that `stargazer` will choke on arithmetic operators in labels like `^`, to avoid this issue you can preface the character with two forward slashes.

```{r, eval=T}
stargazer(m1, 
          type = "text",
          report = "vct*", # show t-statistics too
          star.cutoffs = c(0.05, 0.01, 0.001), digits=5,digits.extra=0,
          covariate.labels=c("Years of schooling","Years of schooling\\^2"),
          dep.var.labels="Income ($10,000s)")
```

**Note that of the three models we ran in the last and this lab, this one is producing the best R^2^.  In the "level-level" model (income~schooling), we had an R^2^ of around .13, in the "level-log" model (income~ln(schooling)) our R^2^ decreased to around .11**.  

## 11. Write the regression equation for this polynomial regression.

### Answer

$$\hat{income10k}=3.47388 + -0.431(educ) +  0.02752(educ^2)$$
*Predicted income equals the intercept of 3.47 plus the linear term of -0.431 times education plus the quadratic term of 0.028 times education squared.*

## 12. Interpret the intercept and slopes

*Remember that acceleration parameters do not have an intuitive interpretation, just say what the sign tells us.*

### Answer

Intercept: Predicted income when someone has zero years of schooling is $34,740.

Slope: When education is zero, a yearly increase in schooling is associated with a decrease in income of $4310.  (not significant)

Acceleration parameter: Because it is positive, this tells us that the association between schooling and income gets more positive as schooling increases.  We could use calculus and multiply the slope by 2 to tell us that a yearly increase in education means the slope for education will become more positive by .056 $10,000 (or $560).

## 13. Show the calculations for the predicted income for someone with 20 years of education

### Answer

$$\hat{income10k}=3.47388 + -0.431(20) +  0.02752(20^2) = 5.8619$$ or $58,619$$.

## 14. Use effects package to produce predictions

Produce all potential levels of predicted income from 4 to 20 years of education in 1 year increments. Make sure your term matches the model term you used. Save these predictions as a data frame all in one line of code.

```{r, eval=T}
eff_df <- as.data.frame(effect("EDUC",
                               m1,
                               xlevels = list(EDUC = seq(4,20,1))))
eff_df
```

**Note the prediction for 20 years--it should be pretty close to the one we calculated.  I calculated a prediction of 5.8619 and `effect` is giving me a prediction of 5.86104.**

## 15 Create a plot

Use the data frame of predicted effects to create a plot. Be sure to include a meaningful title and axis labels.


```{r}
ggplot(data = eff_df, aes(x = EDUC, y = fit*10000)) +
  geom_ribbon(alpha = 0.2, fill = "yellow", color = "yellow",
              aes(ymin = lower*10000, ymax = upper*10000)) +
  geom_point() +
  labs(x = "Years of Schooling", y = "Predicted Income",
       title = "Personal Income By Education") +
  theme_minimal()
```



## 16. Calculate the amount of schooling when the instantaneous effect is equal to 0

This refers to when the relationship between income and education is exactly equal to zero. See lecture notes 14, pages 34-35 for more information. 

### Answer

$$\hat{income10k}=3.47388 + -0.431(educ) +  0.02752(educ^2)$$


$$\frac{dy}{dx}=-.431 + .02752(2)(educ)$$

Deflection point: $$\frac{-\hat{b_1}}{2 \times \hat{b_2}}=\frac{0.431}{.05504}=7.83067$$

The association between schooling and income is negative when schooling is between 0 and 7.83 years, with the deflection point being 7.83 years.  For schooling beyond 7.83 years, the association is positive.  

## 17 In the subsetted data frame, create a logical variable to show deflection point cutoff

Create a logical (`TRUE`/`FALSE`) variable that is equal to TRUE if the year of education for a given person is below the deflection point (step 18). This is the exact point where the relationship between income and years of education reverses (i.e., goes from negative to positive, or vice versa).


```{r}
df_ss$deflection<-df_ss$EDUC < 7.83067
```

## 18 Create a proportion table

Show the proportion of people whose years of schooling is at or below the deflection point and write up this number in a sentence.

```{r, eval=T}
prop.table(table(df$deflection))
```
**(the low, low proportion (.036) of people with such a low level of education suggests that we really should not make much of the negative association or the dramatic U-turn in the association)**

## 19. Save R script


