---
title: "Lab 14"
output: 
  word_document:
    highlight: kate
    reference_docx: style_template2.docx
    fig_caption: no
---

```{r setup, eval=T,include=F}
```

In this assignment, you will use the `ecls.rda` dataset and conduct two simple linear regressions where spring kindergarten reading scores are the outcome.  One predictor is the family’s income, and the other predictor is the number of children less than 18 years old who live in the household.

As you progress through the assignment, you may need to customize `options(scipen=VALUE)` to encourage R to use or not use scientific notation (the higher the value, the less likely R is to use scientific notation) and `options(digits=VALUE)` to encourage R to round to a certain number of digits.


* Integrate your answers in this Word document .  
* For any problem asking you to carry out an operation in R, copy and paste your syntax and any R output below the respective question.
* For questions that ask you to do manual calculations, you should show your work. Using R for manual calculations is allowed and encouraged (show your R syntax and output).
* For each lab assignment, you should maintain an organized organized syntax file ("script")
  + With comments prefaced by `#` that explain what you are doing
  + missteps/mistakes are deleted
  + in case you lose your work, you can run your script and reproduce your analyses.
  + Some labs may require you to build on analyses/data cleaning you did in a prior lab and saving scripts easily allow you to do that
* You need to upload this document to Canvas by the deadline *and check Canvas to make sure your upload took*

# Problem 1 (0 pts)

Preliminaries.  Install and load into memory these packages: `haven` (you will not use any `haven` functions, but you will read in a `haven`-formatted dataframe and so it is good to have the package loaded);  `ggplot2` (to make graphs); `labelled` (to view variable labels); `psych` (we will use the describe function for descriptives); and `stargazer` (to put your regression results into nicely formatted tables).  

# Problem 2 (0 pts)

Load the `ecls.rda` datafile.

# Problem 3 (1 pt)

Ask R to produce the variable labels for all labels.

Syntax:

```{r,eval=F}
attributes(DATAFRAME)["variable.labels"] 
```

# Problem 4 (3 pt)

Create *z*-standardized versions of the reading score, family income, and # of children in the household variables.  Do NOT overwrite the original variables!

Syntax:

```{r, eval=F}
DATAFRAME$NEWVAR<-as.numeric(scale(DATAFRAME$VARIABLE))
```

# Problem 5 (1 pt)

Create a new variable that is family income (which is originally in dollars) rescaled so it is in thousands. 

# Problem 6 (1 pt)

Create a subsetted dataframe containing the childid, reading score (original and z-standardized), income (rescaled and z-standardized), # of kids (original and z-standardized) variables.  The dataframe should only consist of subjects with complete information on these variables.


Syntax:

```{r,eval=F}
NEWDATAFRAME<-subset(ORIGINALDATAFRAME, !is.na(VARIABLE1) & !is.na(VARIABLE2) & ... & !is.na(LASTVARIABLE), select=c(VARIABLE1, VARIABLE2, ..., LASTVARIABLE))
```

# Problem 7 (2 pts)

Get the number of cases in the original ECLS data frame and the number of cases you get after deleting cases with missing values.  WHat is the number of cases dropped?

Syntax:

```{r, eval=F}
nrow(DATAFRAME)
```

# Problem 8 (1 pt)

To be safe, remove the original ecls dataframe from your R workspace.

Syntax:

```{r, eval=F}
rm(DATAFRAME)
```

# Problem 9 (6 pts)

Construct a histogram of the distribution of the test score, income, and # of children variables (use the original variables).  The histograms should show percentages and have clear axis labels. 

Syntax:

```{r, eval=F}
ggplot(DATAFRAME, aes(VARIABLE))+
		geom_histogram(aes(y=(..count..)*100/sum(..count..)))+
		ylab("LABEL TEXT")+
		xlab("LABEL TEXT")
```

# Problem 10 (2 pts)

Run separate bivariate regressions of test score (dependent variable) on family income (in $1,000s) and number of children in the household.  Do NOT use the *z*-standardized versions of the variables.  Store the results in separate objects.

Syntax:

```{r, eval=F}
MODELOBJECT<-lm(DEPENDENTVARIABLE~INDEPENDENTVARIABLE, data=DATAFRAME)
```

# Problem 11 (5 pts)

Use `stargazer` to present the results for both models in one table.  Your table should have clear labels for the dependent variable and the predictors.  The units of measurement of both the predictors and outcome should be clear.  You should NOT use the default presentation of a single star representing *marginal* (p<.10) significance.

Here is some `stargazer` syntax you may find helpful.  I recommend running stargazer using just the `type="text"` option and then adding the options one at a time to see how they affect your table.  You should be especially careful with `covariate.labels` as you do NOT want to mislabel your predictors.  If you have questions about the `stargazer` syntax the [documentation](https://cran.r-project.org/web/packages/stargazer/stargazer.pdf) is helpful.  

Syntax:

```{r,eval=F}
stargazer(MODELOBJECT,MODELOBJECT,
          type="text",
          report="vcs*",
          title="text",
          dep.var.caption="TEXT",
          dep.var.labels=c("TEXT1","TEXT2"),
          covariate.labels=c("TEXT1","TEXT2",...,"TEXTLAST"),
          star.cutoffs=c(VALUE1,VALUE2,VALUE3),
          star.char=c("SYMBOL1","SYMBOL2","SYMBOL3"),
          notes="TEXT",
          notes.append=FALSE
          )

```

# Problemn 12 (4 pts)

Give a textbook interpretation of the intercept in both models.  Your interpretation should make clear the units of measurement.

# Problem 13 (4 pts)

Give a textbook interpretation of the slope in both models.  Your interpretation should make clear the units of measurement.

# Problem 14 (4 pts)

Give a textbook interpretation of the r^2^ in both models.

# Problem 15 (2 pts)

Re-do the models, but this time make all variables be *z*-standardized.

# Problem 16 (2 pt)

Re-do your `stargazer` table with the new models.  Your table should make clear the units of measurement (SD).

# Problem 17 (2 pts)

Give a textbook interpretation for the intercept for the model where # of children in the household is the predictor.

# Problem 18 (2 pts)

Give a textbook interpretation for the slope for the model where # of children in the household is the predictor.

# Problem 19 (2 pts)

Ask R to report summaries of the model objects so you can get the *p* values for the slopes. (it does not matter if you use a model using the raw metrics or the *z*-standardized variables).

Syntax:

```{r,eval=F}
summary(MODELOBJECT)
```

# Problem 20 (6 pts)

In ONE sentence, write-up the results of both regressions in APA style.

# Problem 21 (0 pts)

Upload your lab to Canvas.  Double check to make sure the upload took.
