---
title: "Lab 8"
author: "Joshua Klugman"
output:
  word_document:
    reference_docx: "../lab_08.docx"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
root_dir <- normalizePath("..")
knitr::opts_knit$set(root.dir = root_dir)
.libPaths(c(file.path(root_dir, "rlib"), .libPaths()))

library(tidyverse)
library(haven)
library(emmeans)
library(afex)
library(pastecs)
library(car)
source(file.path(root_dir, "Rallfun-v45.txt"))
```

# Lab 8 overview

The data in the file `wordsum v2.sav` is a vocabulary test administered to people from different regions of the United States, with scores ranging from 0 to 10. We have a 2×4 design, with our two factors being a city resident (versus not being a city resident; this is variable `sized`) and region (Midwest, Northeast, South, and West; this is variable `regions`). The outcome is the vocabulary score (variable `ws` which is a count variable ranging from 0 correct words to 10 correct words).

## Problem 1 (0 pts)

Load packages and Wilcox functions.

Syntax:

```{r p1_syntax, eval=FALSE}
library(afex)
library(car)
library(emmeans)
library(haven)
source("Rallfun-v45.txt")
```

Output:

```{r p1_output, echo=FALSE}
library(afex)
library(car)
library(emmeans)
library(haven)
source("Rallfun-v45.txt")
```

## Problem 2 (0 pts)

Read in the `wordsum v2.sav` datafile.

Syntax:

```{r p2_syntax, eval=FALSE}
ws <- read_sav("wordsum v2.sav")
head(ws)
```

Output:

```{r p2_output, echo=FALSE}
ws <- read_sav("wordsum v2.sav")
head(ws)
```

## Problem 3 (6 pts)

Look at the value labels for `sized` and `regions`, create factor versions, and verify construction.

Syntax:

```{r p3_syntax, eval=FALSE}
attr(ws$sized, "labels")
attr(ws$regions, "labels")

ws$sized_f <- factor(ws$sized,
                     levels = c(0, 1),
                     labels = c("Non-City", "City"))

ws$regions_f <- factor(ws$regions,
                       levels = c(1, 2, 3, 4),
                       labels = c("Midwest", "Northeast", "South", "West"))

table(ws$sized, ws$sized_f, exclude = NULL)

table(ws$regions, ws$regions_f, exclude = NULL)
```

Output:

```{r p3_output, echo=FALSE}
attr(ws$sized, "labels")
attr(ws$regions, "labels")

ws$sized_f <- factor(ws$sized,
                     levels = c(0, 1),
                     labels = c("Non-City", "City"))

ws$regions_f <- factor(ws$regions,
                       levels = c(1, 2, 3, 4),
                       labels = c("Midwest", "Northeast", "South", "West"))

table(ws$sized, ws$sized_f, exclude = NULL)

table(ws$regions, ws$regions_f, exclude = NULL)
```

## Problem 4 (4 pts)

Create a superfactor based on the city resident and region variables, and cross-tabulate.

Syntax:

```{r p4_syntax, eval=FALSE}
ws <- group_by(ws, sized_f, regions_f)
ws <- mutate(ws, regionsize_new = cur_group_id())
ws <- ungroup(ws)

ws$regionsize_new_f <- factor(ws$regionsize_new,
                              levels = 1:8,
                              labels = c("Midwest Non-City", "Northeast Non-City",
                                         "South Non-City", "West Non-City",
                                         "Midwest City", "Northeast City",
                                         "South City", "West City"))

table(ws$regionsize_new_f, ws$sized_f, exclude = NULL)

table(ws$regionsize_new_f, ws$regions_f, exclude = NULL)
```

Output:

```{r p4_output, echo=FALSE}
ws <- group_by(ws, sized_f, regions_f)
ws <- mutate(ws, regionsize_new = cur_group_id())
ws <- ungroup(ws)

ws$regionsize_new_f <- factor(ws$regionsize_new,
                              levels = 1:8,
                              labels = c("Midwest Non-City", "Northeast Non-City",
                                         "South Non-City", "West Non-City",
                                         "Midwest City", "Northeast City",
                                         "South City", "West City"))

table(ws$regionsize_new_f, ws$sized_f, exclude = NULL)

table(ws$regionsize_new_f, ws$regions_f, exclude = NULL)
```

## Problem 5 (2 pts)

Produce histograms for each cell.

Syntax:

```{r p5_syntax, eval=FALSE}
ggplot(data = ws, aes(x = ws)) +
  geom_histogram() +
  facet_grid(regions_f ~ sized_f)
```

Output:

```{r p5_output, echo=FALSE, fig.cap="Vocabulary scores by region and city status"}
ggplot(data = ws, aes(x = ws)) +
  geom_histogram() +
  facet_grid(regions_f ~ sized_f)
```

## Problem 6 (1 pt)

Skewness statistics and Shapiro-Wilk tests for each cell.

Syntax:

```{r p6_syntax, eval=FALSE}
tapply(ws$ws, ws$regionsize_new_f, pastecs::stat.desc, norm = TRUE)
```

Output:

```{r p6_output, echo=FALSE}
tapply(ws$ws, ws$regionsize_new_f, pastecs::stat.desc, norm = TRUE)
```

## Problem 7 (1 pt)

Levene's test for equal variances.

Syntax:

```{r p7_syntax, eval=FALSE}
car::leveneTest(ws ~ regionsize_new_f, data = ws, center = mean)
```

Output:

```{r p7_output, echo=FALSE}
car::leveneTest(ws ~ regionsize_new_f, data = ws, center = mean)
```

## Problem 8 (3 pts)

Use the MAD-Median rule to detect outliers.

Syntax:

```{r p8_syntax, eval=FALSE}
o <- tapply(ws$ws, ws$regionsize_new_f, outpro)
o
# If outliers appear in any group, run this for each group with outliers:
# ws$ws[ws$regionsize_new_f=="GROUP NAME"][o[["GROUP NAME"]]$out.id]
```

Output:

```{r p8_output, echo=FALSE}
o <- tapply(ws$ws, ws$regionsize_new_f, outpro)
o
```

## Problem 9 (5 pts)

Diagnostics write-up (3 sentences maximum).

Based on the histograms and skewness statistics, the distributions appear only modestly skewed across cells. Shapiro-Wilk tests indicate statistically significant departures from normality in each cell (all p-values < .01). Levene’s test suggests homogeneity of variances across the eight cells (p = .8907), and the MAD-Median rule indicates no outliers in any cell.

## Problem 10 (1 pt)

Create an ID variable.

Syntax:

```{r p10_syntax, eval=FALSE}
ws$id <- rownames(ws)
```

Output:

```{r p10_output, echo=FALSE}
ws$id <- rownames(ws)
```

## Problem 11 (3 pts)

Produce cell and marginal means.

Syntax:

```{r p11_syntax, eval=FALSE}
ws_anova <- aov_car(ws ~ sized_f * regions_f + Error(id), data = ws)

emmeans(ws_anova, ~ sized_f * regions_f)

emmeans(ws_anova, ~ sized_f)
emmeans(ws_anova, ~ regions_f)
```

Output:

```{r p11_output, echo=FALSE}
ws_anova <- aov_car(ws ~ sized_f * regions_f + Error(id), data = ws)

emmeans(ws_anova, ~ sized_f * regions_f)

emmeans(ws_anova, ~ sized_f)
emmeans(ws_anova, ~ regions_f)
```

## Problem 12 (5 pts)

Excel table of cell means.

Syntax:

```{r p12_syntax, eval=FALSE}
cell_means <- as.data.frame(emmeans(ws_anova, ~ regions_f * sized_f))
cell_table <- cell_means |>
  dplyr::select(regions_f, sized_f, emmean) |>
  pivot_wider(names_from = sized_f, values_from = emmean)
cell_table
```

Output:

```{r p12_output, echo=FALSE}
cell_means <- as.data.frame(emmeans(ws_anova, ~ regions_f * sized_f))
cell_table <- cell_means |>
  dplyr::select(regions_f, sized_f, emmean) |>
  pivot_wider(names_from = sized_f, values_from = emmean)
cell_table
```

Create a 4x2 table in Excel with rows as `regions_f` and columns as `sized_f`. Use the cell means above, round to two decimals, and format with a clear title and minimal borders.

## Problem 13 (4 pts)

Explain whether it is safe to interpret the main effects.

The interaction between city status and region is significant, so the main effects are not safe to interpret. The city vs. non-city differences are not consistent across regions, and the rank order of regions changes across city status.

## Problem 14 (1 pt)

Two-way ANOVA output.

Syntax:

```{r p14_syntax, eval=FALSE}
summary(ws_anova)
ws_anova$Anova
```

Output:

```{r p14_output, echo=FALSE}
summary(ws_anova)
ws_anova$Anova
```

## Problem 15 (6 pts)

Calculate partial omega squared for each effect.

Syntax:

```{r p15_syntax, eval=FALSE}
an <- as.data.frame(ws_anova$Anova)
ss <- an$`Sum Sq`
df <- an$Df
names(ss) <- row.names(an)
names(df) <- row.names(an)

ss_error <- ss["Residuals"]
df_error <- df["Residuals"]
mse <- ss_error / df_error

effects <- setdiff(row.names(an), c("(Intercept)", "Residuals"))
ss_effect <- ss[effects]
df_effect <- df[effects]

omega_p <- (ss_effect - df_effect * mse) / (ss_effect + ss_error + mse)

omega_table <- data.frame(
  effect = effects,
  df = df_effect,
  SS = ss_effect,
  SS_error = ss_error,
  MSE = mse,
  omega_p = omega_p
)
omega_table
```

Output:

```{r p15_output, echo=FALSE}
an <- as.data.frame(ws_anova$Anova)
ss <- an$`Sum Sq`
df <- an$Df
names(ss) <- row.names(an)
names(df) <- row.names(an)

ss_error <- ss["Residuals"]
df_error <- df["Residuals"]
mse <- ss_error / df_error

effects <- setdiff(row.names(an), c("(Intercept)", "Residuals"))
ss_effect <- ss[effects]
df_effect <- df[effects]

omega_p <- (ss_effect - df_effect * mse) / (ss_effect + ss_error + mse)

omega_table <- data.frame(
  effect = effects,
  df = df_effect,
  SS = ss_effect,
  SS_error = ss_error,
  MSE = mse,
  omega_p = omega_p
)
omega_table
```

## Problem 16 (5 pts)

Write up the results of the omnibus test.

A two-way ANOVA showed a significant main effect of city status, F(1, 312) = 6.19, p = .013, ω²p = 0.016, and a significant main effect of region, F(3, 312) = 3.03, p = .030, ω²p = 0.019. There was also a significant city status × region interaction, F(3, 312) = 18.68, p < .001, ω²p = 0.144. Because the interaction is significant, interpretation should focus on the cell means rather than the marginal main effects.

## Problem 17 (0 pts)

Upload this document to Canvas and verify the submission.
